---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval=F)
```

# R Basic Data Analysis Primer and Cheat Sheet

## Coding Variables

Make sure your variables are of the correct type before doing analysis.

Categorical variables should be set as _factors_ and continuous variables set as _numerics_

```{r}

#Set a variable as a factor:
dataFrame$variable<-as.factor(dataFrame$variable)

#Set a variable as a numeric:
dataFrame$variable<-as.numeric(dataFrame$variable)

```

***

## Summary Statistics

**Mean:**
```{r}

mean(dataFrame$x,na.rm=T)

```

**Median:**
```{r}

median(dataFrame$x,na.rm=T)

```

**Standard Deviation:**
```{r}

sd(dataFrame$x,na.rm=T)

```

**Summary:**
```{r}

#This function outputs a number of useful summary values
summary(dataFrame$x)

```

***

## Difference Between Two Means

**Between-Subjects _t_-Test**

To test the difference between the means of two different groups of people, do a between-subjects _t_-test:

```{r}

tTestName<-t.test(outcome~predictor,
                  data=dataFrame)
#Outcome = your outcome variable
#Predictor = a two-level predictor variable
#dataFrame = the dataframe containing your data

#To get the output of your t-test, execute:
tTestName

```

**Within-Subjects _t_-Test**

To test the difference between the means of two measurements of the same people at two different times, do a within-subjects _t_-test:

```{r}

tTestName <- t.test(dataFrame$measure2, 
                    dataFrame$measure1, 
                    paired = T)

#measure2 = your second measure of each person
#measure1 = your first measure of each person
#"paired = T" = do a within-subjects t.test
#dataFrame = the dataframe containing your data

#To get the output of your t-test, execute:
tTestName

```

***

## Difference Across Multiple Means

To test the difference between the means of 3 or more groups of people that differ along a single variable, do a one-way ANOVA:

```{r}

anovaName<-lm(outcome~predictor,
              data=dataFrame)

#outcome = your outcome variable
#predictor = a categorical predictor variable with at least 3 levels
#dataFrame = the dataframe containing your data

#To get the output of your ANOVA, execute:
anova(anovaName)

#To view means as a function of group, do:

tapply(dataFrame$outcome,
       dataFrame$predictor,
       function(x) 
         mean(x,na.rm=T))

```

To test the difference between group means as a function of multiple variables, do an _n_-way ANOVA:

```{r}

anovaName<-lm(outcome~predictor1*predictor2,
              data=dataFrame)

#outcome = your outcome variable
#predictor1 = your first categorical predictor variable
#predictor2 = your second categorical predictor variable
#dataFrame = the dataframe containing your data

#To get the output of your ANOVA, execute:
anova(anovaName)

#To view means as a function of the predictors, do:

tapply(dataFrame$outcome,
       list(dataFrame$predictor1,
            dataFrame$predictor2),
       function(x)
         mean(x,na.rm=T))

```

To test the difference between multiple (>=3) means measured on the same people at different points in time, do a within-subjects ANOVA.

To do this, you first must "melt" your data from "wide format" to "long format."

Often we store data like this:

Participant | Observation 1 | Observation 2
------------|---------------|--------------
1           | 4             | 3
2           | 7             | 1

This is "wide" data. Melting data reorganizes it to "long" format like this:

Participant | Observation # | Score
------------|---------------|------
1           | 1             | 4
1           | 2             | 3
2           | 1             | 7
2           | 2             | 1

One easy way to do this is the _melt()_ function inside the **reshape2** package:

```{r}
#Make sure you run library(reshape2) before you try this

dataMelted<-melt(dataFrame,
                 measure.vars=c("measure1","measure2"),
                 variable.name="observationNumber",
                 value.name="score")

#dataFrame = the dataframe you want melted
#"measure1" = your first measure of the outcome variable
#"measure2" = your second measure of the outcome variable
#variable.name = what you want your observation # variable to be called
#value.name = what you wnat your score variable to be called
```

You can then do a a within-subjects ANOVA on the melted data:

```{r}

anovaName<-aov(outcome~predictor1*predictor2+Error(PIN),
               data=dataFrame)

#outcome = your outcome variable
#predictor1 = your first predictor variable
#predictor 2 = your second predictor variable (optional)
#PIN = a unique participant identifier
#dataFrame = the dataframe containing your data

```

***

## Associations Between Continuous Variables

If you want to see if two continuous variables are associated with one another, do a correlation test:

```{r}

corrName<-cor.test(dataFrame$x,
                   dataFrame$y)

#x = your first continuous variable
#y = your second continuous varaible
#dataFrame = the dataframe containing your data

#To get the output of your correlation, execute:
corrName

```

If you want to see if one continuous outcome variable is associated with multiple predictors (continuous or categorical), do a regression:

```{r}

#For just main effects:
regName<-lm(outcome~predictor1+predictor2,
            data=dataFrame)

#For interactions:
regName<-lm(outcome~predictor1*predictor2,
            data=dataFrame)

#outcome = your outcome variable
#predictor1 = your first predictor variable, categorical or continuous
#predictor2 = your second predictor variable, categorical or continuous
#dataFrame = the dataframe containing your data

#To get the output of your regression, execute:
summary(r1)

```


\newpage
# Basic Data Cheat Sheet

Test                   | Code 
-----------------------|--------------------------------------------------------------------
Mean                   | mean(dataFrame\$x, na.rm=T), colMeans(dataFrame[ ,_$$n_1$$_:_$$n_2$$_], na.rm=T)
Median                 | median(dataFrame\$x, na.rm=T)
Standard deviation     | sd(dataFrame\$x, na.rm=T)
Summary                | summary(dataFrame\$x)
-----------------------|--------------------------------------------------------------------
Between-subjects t-test| tTestName<-t.test(y~x, data=dataFrame)
Within-subjects t-test | tTestName<-t.test(dataFrame\$measure1, dataFrame\$measure2, paired=T)
-----------------------|--------------------------------------------------------------------
One-way ANOVA          | anovaName<-lm(y~x, data=dataFrame)
Two-way ANOVA          | anovaName<-lm(y~x1*x2, data=dataFrame)
Melt data (reshape2)   | dataMelt<-melt(dataFrame, measure.vars=c("y1","y2"), variable.name="x", value.name="y")
Within-subjects ANOVA  | anovaName<-aov(y~x+Error(PIN),  data=dataMelt)
-----------------------|--------------------------------------------------------------------
Correlation            | cor.test(dataFrame\$x, dataFrame\$y)
Regression             | regName<-lm(y~x1+x2, data=dataFrame)
