---
title: "Data Analysis of Week 3 Example Data"
author: "Daniel Conroy-Beam"
date: "`r Sys.Date()`"
output:
   html_document:
    toc: true
    toc_float: true
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(reshape2)
library(ggplot2)

```


## Overview

This report presents primary data analysis for the Week 3 example data. Specifically, this will report basic descriptive statistics as well as report tests  of the following predictions:  

1. People will feel more guilt when:  
  + 1.1. They caused the shock (vs. their partner)
  + 1.2. The partner is morally good (vs. bad)
  + 1.3. The shock is stronger (high > medium > low)
2. Women will report more guilt than men
3. Agreeable people will report more guilt
4. Neurotic people will report more guilt
  
and  
  
5. People will be more generous when:
  + 5.1 They caused the shock (vs. their partner)
  + 5.2 The partner is morally good (vs. bad)
  + 5.3 The shock is stronger (high > medium > low)
  
6. Women will be more generous than men  
7. Agreeable people will be more generous  
8. Neurotic people will be more generous  



```{r load}

#Load in the data
data<-read.csv("Week 3/Processed Data/Week 3 Data PROCESSED.csv")

#Recode variables:
data$PIN<-as.factor(data$PIN)
data$sex<-as.factor(data$sex)
data$shock<-as.factor(data$shock)
data$shockCause<-as.factor(data$shockCause)
data$pMoral<-as.factor(data$pMoral)

```



## Descriptives

Descriptive statistics for guilt:
```{r descGuilt}

c(summary(data$guilt),"SD"=sd(data$guilt,na.rm=T))

```

Descriptive statistics for generosity:
```{r descGen}

#Pre-shock:
c(summary(data$preShare),"SD"=sd(data$preShare,na.rm=T))

#Post-shock:
c(summary(data$postShare),"SD"=sd(data$postShare,na.rm=T))

```

Descriptive statistics for personality:
```{r descPers}

#Agreeableness:
c(summary(data$agre),"SD"=sd(data$agre,na.rm=T))

#Neuroticism:
#Pre-shock:
c(summary(data$neuro),"SD"=sd(data$neuro,na.rm=T))

```



## Guilt Analyses

### Prediction 1.1

Prediction 1.1 was supported. People felt more guilty when they caused the shock:
```{r p1.1}

#Run the t-test for prediction 1.1
p1.1<-t.test(data$guilt~data$shockCause)

#Output the result
p1.1

#Generate the graph
plot1.1 <- ggplot(data = data,
                  aes(x = shockCause,
                      y = guilt,
                      fill = shockCause)) +
  labs(x = "Cause of Shock", 
       y = "Guilt") +
  stat_summary(fun = "mean",
               geom = "bar",
               position = "dodge") +
  stat_summary(fun.data = "mean_cl_boot",
               geom = "errorbar",
               position = "dodge") +
  scale_x_discrete(labels = c("Partner", "Self")) +
  theme_gray(base_size = 20) +
  guides(fill=F)

#Output the graph
plot1.1

```



### Prediction 1.2 

Prediction 1.2 was supported. People felt more guilty when the partner was morally good:
```{r p1.2}

#Run the t-test for prediction 1.2
p1.2<-t.test(data$guilt~data$pMoral)

#Output the result
p1.2

#Generate the graph
plot1.2 <- ggplot(data = data,
                  aes(x = pMoral,
                      y = guilt,
                      fill = pMoral)) +
  labs(x = "Partner Morality", y = "Guilt") +
  geom_boxplot() +
  scale_x_discrete(labels = c("Morally Bad", "Morally Good")) +
  theme_gray(base_size = 20) +
  guides(fill=F)

#Output the graph
plot1.2

```



### Prediction 1.3

Prediction 1.3 was supported. People felt more guilty when the shock was stronger:
```{r p1.3}

#Run the ANOVA for prediction 1.3
p1.3<-lm(guilt~shock,data=data)

#Output the result
anova(p1.3)

#Output the group means
tapply(data$guilt,data$shock,function(x) mean(x,na.rm=T))


#Generate the graph
plot1.3 <- ggplot(data = data,
                  aes(x = shock,
                      y = guilt,
                      fill = shock)) +
  labs(x = "Shock Voltage", y = "Guilt") +
  geom_violin(position = position_dodge(.9)) +
  geom_boxplot(position = position_dodge(.9),width=.1) +
  scale_x_discrete(labels = c("Low", "Medium", "High")) +
  theme_gray(base_size = 20) +
  guides(fill=F)

#Output the graph
plot1.3

```



### Prediction 2

Prediction 2 was not supported. Women did not report more guilt than men:
```{r p2}

#Run the t-test for prediction 2
p2<-t.test(data$guilt~data$sex)

#Output the result
p2

#Generate the graph
plot2 <- ggplot(data = data,
                  aes(x = sex,
                      y = guilt,
                      fill = sex)) +
  labs(x = "Participant Sex", y = "Guilt") +
  geom_violin(position = position_dodge(.9)) +
  geom_boxplot(position = position_dodge(.9),width=.1) +
  scale_x_discrete(labels = c("Female","Male")) +
  theme_gray(base_size = 20) +
  guides(fill=F)

#Output the graph
plot2

```



### Prediction 3

Prediction 3 was supported. Agreeable people reported more guilt:
```{r p3}

#Run the correlation test for prediction 3
p3<-cor.test(data$agre,data$guilt)

#Output the result
p3

#Generate the plot
plot3<-ggplot(data=data,
              aes(x=agre,
                  y=guilt))+
  labs(x="Participant Agreeableness",
       y="Guilt")+
  geom_point()+
  geom_smooth(method="lm")+
  theme_gray(base_size=20)

#Output the plot
plot3

```



### Prediction 4

Prediction 4 was not supported. There was no relationship between neuroticism and guilt:
```{r p4}

#Run the correlation test for prediction 4
p4<-cor.test(data$neuro,data$guilt)

#Output the result
p4

#Test predictions 3 and 4 in a regression:
guiltReg<-lm(guilt~neuro+agre,data=data)
summary(guiltReg)

#Generate the plot
plot4<-ggplot(data=data,
              aes(x=neuro,
                  y=guilt))+
  labs(x="Participant Neuroticism",
       y="Guilt")+
  geom_jitter()+
  geom_smooth(method="lm")+
  theme_gray(base_size=20)

#Output the plot
plot4

```



## Generosity Analyses

```{r melt}

#Melt the data for within-subjects analyses
dataMelt<-melt(data,
               measure.vars=c("preShare","postShare"),
               variable.name=c("prePost"),
               value.name=c("share"))

```



### Prediction 5.1

Prediction 5.1 was supported. People were more generous when they caused the shock:
```{r p5.1}

#Run the within-subjects ANOVA for p5.1
p5.1<-aov(share~prePost*shockCause+Error(PIN),data=dataMelt)

#Output the results:
summary(p5.1)

#Output means as a function of condition:
tapply(dataMelt$share,dataMelt[,c(23,20)],function(x) mean(x,na.rm=T))

#Generate the plot
plot5.1<-ggplot(data = dataMelt,
                aes(x = shockCause,
                    y = share,
                    fill = prePost)) +
  geom_violin(position = position_dodge(.9)) +
  geom_boxplot(position = position_dodge(.9), width = .1) +
  labs(x = "Cause of Shock", y = "Money Shared ($)", fill = "Time") +
  scale_x_discrete(labels = c("Partner", "Self")) +
  scale_fill_discrete(labels = c("Before Shock", "After Shock")) +
  theme_grey(base_size = 20)

#Output the plot
plot5.1
```



### Prediction 5.2 

Prediction 5.2 was supported. People were more generous when the partner was morally good:
```{r p5.2}

#Run the within-subjects ANOVA for p5.2
p5.2<-aov(share~prePost*pMoral+Error(PIN),data=dataMelt)

#Output the results:
summary(p5.2)

#Output means as a function of condition:
tapply(dataMelt$share,dataMelt[,c(23,21)],function(x) mean(x,na.rm=T))

#Generate the plot
plot5.2<-ggplot(data = dataMelt,
                aes(x = pMoral,
                    y = share,
                    fill = prePost)) +
  geom_violin(position = position_dodge(.9)) +
  geom_boxplot(position = position_dodge(.9), width = .1) +
  labs(x = "Partner Morality", y = "Money Shared ($)", fill = "Time") +
  scale_x_discrete(labels = c("Morally Bad", "Morally Good")) +
  scale_fill_discrete(labels = c("Before Shock", "After Shock")) +
  theme_grey(base_size = 20)

#Output the plot
plot5.2
```



### Prediction 5.3

Prediction 5.3 was supported. People were more generous when the shock was stronger:
```{r p5.3}

#Run the within-subjects ANOVA for p5.3
p5.3<-aov(share~prePost*shock+Error(PIN),data=dataMelt)

#Output the results:
summary(p5.3)

#Output means as a function of condition:
tapply(dataMelt$share,dataMelt[,c(23,19)],function(x) mean(x,na.rm=T))

#Generate the plot
plot5.3 <- ggplot(data = dataMelt,
                  aes(x = shock,
                      y = share,
                      fill = prePost)) +
  geom_violin(position = position_dodge(.9)) +
  geom_boxplot(position = position_dodge(.9), width = .1) +
  labs(x = "Shock Voltage", y = "Money Shared ($)", fill = "Time") +
  scale_x_discrete(labels = c("Low", "Medium", "High")) +
  scale_fill_discrete(labels = c("Before Shock", "After Shock")) +
  theme_grey(base_size = 20)

#Output the plot
plot5.3
```



### Prediction 6

Prediction 6 was not supported. Women were not more generous:
```{r p6}

#Run the ANOVA to test prediction 6
p6<-aov(share~prePost*sex+Error(PIN),data=dataMelt)

#Output the results:
summary(p6)

#Output means as a function of condition and sex:
tapply(dataMelt$share,dataMelt[,c(23,2)],function(x) mean(x,na.rm=T))

#Generate the plot
plot6<-ggplot(data = dataMelt,
                aes(x = sex,
                    y = share,
                    fill = prePost)) +
  geom_violin(position = position_dodge(.9)) +
  geom_boxplot(position = position_dodge(.9), width = .1) +
  labs(x = "Participant Sex", y = "Money Shared ($)", fill = "Time") +
  scale_x_discrete(labels = c("Female", "Male")) +
  scale_fill_discrete(labels = c("Before Shock", "After Shock")) +
  theme_grey(base_size = 20)

#Output the plot
plot6

```



### Prediction 7

Prediction 7 was supported. Agreeable people were more generous:
```{r p7}

#Run the correlation test for prediction 7
p7<-cor.test(data$agre,data$postShare)

#Output the result
p7


#Generate the plot
plot7<-ggplot(data=dataMelt,
              aes(x=agre,
                  y=share,
                  color=prePost))+
  geom_jitter(alpha=.2)+
  geom_smooth(method="lm",linewidth=2)+
  labs(x="Participant Agreeableness",
       y="Money Shared ($)",
       color="Time")+
  scale_color_discrete(labels=c("Pre-Shock","Post-Shock"))+
  theme_gray(base_size=20)

#Output the plot
plot7
```



### Prediction 8

Prediction 8 was not supported. There was no relationship between neuroticism and generosity:
```{r p8}

#Run the correlation test for prediction 8
p8<-cor.test(data$neuro,data$postShare)

#Output the result
p8

#Test predictions 7 and 8 in a regression:
genReg<-lm(postShare~neuro+agre,data=data)
summary(genReg)

#Generate the plot
plot8<-ggplot(data=dataMelt,
              aes(x=neuro,
                  y=share,
                  color=prePost))+
  geom_jitter(alpha=.2)+
  geom_smooth(method="lm",linewidth=2)+
  labs(x="Participant Neuroticism",
       y="Money Shared ($)",
       color="Time")+
  scale_color_discrete(labels=c("Pre-Shock","Post-Shock"))+
  theme_gray(base_size=20)

#Output the plot
plot8
```
