---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval=F)
```

# R Primer and Cheat Sheet

This document is intended to introduce you to the basics of base R. This does not cover all of R programming, but does cover enough to allow basic data manipulation and analysis. This should teach you enough R to make it easier to learn the rest. This document (particularly the last section) can also serve as a reference guide as you continue to learn R in the future.
	
***

## R in a nutshell:

R uses functions to manipulate data structures.  

You supply R with the data structure you want manipulated and the function you want R to use and R happily obliges.

***

## Data Structures

Data structures are “containers” for your data.  

Three types of data structures that will be highly relevant to you:  

Scalars: holds a single number. “1”, “2”, “3”  
Vectors: holds lists of scalars. “1, 2, 3, 4, 5, 6, 7, 8, 9, 10”  
Dataframes: holds lists of vectors (like a spreadsheet)  

\hfill\break

To put data into a data structure, use <- like so:  

```{r}
structureName <- data
```


You can call your data structures whatever you like. R doesn’t care. It just can’t start with a number or contain certain symbols (e.g. %)  
\hfill\break
Some tips: make your structure names memorable, easy to type, and informative
Try to follow a convention. Some people use snake_casing. I usually use camelCasing.

\hfill\break

To create a scalar, do:
```{r}
scalar1 <- 1
scalar2 <- 2
scalar3 <- 109387958125096128995408

```

To create a vector, use the c() command:
```{r}
vector1 <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
```

You can also use 1:10 to tell R "Give me all of the numbers between 1 and 10:
```{r}
vector2 <- 1:10
vector3 <- 25:34
```

Or, you can use c() to combine data you already have in scalar data structures:
```{r}
vector4 <- c(scalar1, scalar2, scalar3)

```

To create a dataframe, use the data.frame() command:
```{r}
yourData<-data.frame(vector1,vector2,vector3)
```

For this to work, your vectors need to be "of the same length” (i.e., have the same number of scalars in them) or the longer vectors need to be multiples of the shorter vectors (e.g., vectors of length 4 and 2 can be combined, but not vectors of length 4 and 3)

\hfill\break

To change specific elements of your data structures, you use “indexing”.  
In R, indexing is done with square brackets: [ ]  
yourData[row,column] pulls out the specified row and column of yourData.

To change the item in the 5th row and 32nd column, you would do:
```{r}
yourData[5, 32] <- newData
```

To refer to all columns or all rows of a dataframe, you just leave the column or row section of the index empty.
So every column in the 3rd row becomes:
```{r}
yourData[3, ]

```

And every row in the 50th column becomes:
```{r}
yourData[, 50]
```

***

When using R for data analysis, you will most often store your data in a dataframe, with each of your variables represented as a vector within this dataframe.  

There are lots of ways to import data into R. We’ll talk about some more later. But for now, an easy way is to combine the read.csv() and file.choose() commands:  

```{r}
yourData <- read.csv(file.choose())  
```

This will open a new explorer/finder window on your computer and let you navigate to your file. When you select it, it will automatically be imported into R and stored as a dataframe in your global environment.  

Make sure the file you’re importing is .csv format

***

## R and Functions

In R, you ask R to perform different functions on your data structures.

The basic structure of R functions is as follows:

```{r}
functionName(dataStructure,options)
```

**functionName** would be the name of your given function.  
**dataStructure** would be the name of the data structure you want to manipulate.

This will almost always be a vector. Isolated vectors can be referred to just by their name. However, to refer to a vector that is stored within a dataframe, you must use a $ as follows:
```{r}
yourData$vectorName
```

"options" refer to any additional options you want from that function. You can see what options are available for a function by using ? as follows:
```{r}
?functionName
```

This will pull up the help dialog for that function, including a description of all available options.

***

## Common R Functions and Operators

Below are commonly used functions and operators in R, organized into various sections. Just type this code into R and replace things like “x” and “y” with your data structures. 

***

## Basic Arithmetic

These operators do basic arithmetic calculations on your data.

*Addition:*
```{r}
3 + 2
10190835 + 91730513
```

*Subtraction:*
```{r}
400 - 300
0 - 12
```

*Multiplication:*
```{r}
100 * 100
.5 * 5
```

*Division:*
```{r}
1200 / 600
0 / 40
```

*Exponents:*
```{r}
2 ^ 2
5 ^ 2.5
```

*Square root:*
```{r}
sqrt(4)

#Or

4 ^ .5
```

*Summation:*
```{r}
sum(c(1, 2, 3, 4, 5))
```

*Sum all columns or rows:*
```{r}
colSums(dataframe)
rowSums(dataframe)
```

***

## Comparisons and Logicals

These operators compare data structures or evaluate logical statements, but do not necessarily calculate new values. They generally return either “TRUE” if the statement is true and “FALSE” if it is not.  

Handy tip: R treats “TRUE” as being equivalent to 1 and “FALSE” as being equivalent to 0. So TRUE + TRUE = 2; TRUE + FALSE = 1; and FALSE + FALSE = 0

*Equivalence:*
```{r}
4 == 4
4 == 1
4 == c(1, 2, 3, 4)
```

*Non-Equivalence:*
```{r}
4 != 4
4 != 1
```

*Greater or Less Than:*
```{r}
4 > 4
4 < 1
```

*Greater/Less Than or Equal to:*
```{r}
4 >= 4
4 <= 1
```

*Containment:*
```{r}
4 %in% c(1, 2, 3, 4)
4 %in% c(5, 6, 7, 8)
```

*AND:*
```{r}
#Are both of these statements true?
4 > 3 & 4 < 5
4 < 1 & 4 < 7
```

*OR:*
```{r}
#Is at least one of these statements true?
4 < 1 | 4 < 7
4 < 0 | 4 > 100
```

***

## Viewing Your Data

In order to work with your data, it’s often helpful to be able to look at it; both to inspect the individual cases for issues (e.g. missing data), and to see what your variables are actually called.  

In RStudio, you have two easy options to do this. First, you can click the table icon next to your dataframe in the “Environment” tab.  

This is good if you really need to be able to look at the observations. If you just need to remember the name of a variable, you can use the names() function:
```{r}
names(yourData)
```

This will show you all of the vector names in yourdata in order. This is useful and you should use it often.

***

## Summary Statistics

These functions generally take a vector as an input and each compute some common summary statistic value.

*Sample Size:*

For sample size, you have two options. Which you use depends on what kind of data structure you are using. 

For dataframes:
```{r}
nrow(dataframe)
```

For vectors:
```{r}
length(vector)
```

*Mean:*
```{r}
mean(x)

#Note: R really hates missing data. 
#If you have any missing values in your vector, this will return “NA”. 
#To avoid this, use the “na.rm=TRUE” option
#This means “Remove any missing values before you calculate the mean”.
#You can use this for all summary functions.

mean(x,na.rm=T)

```

*Mean of all columns or rows:*
```{r}
colMeans(x)
rowMeans(x)
```

*Median:*
```{r}
median(x)
```

*Standard Deviation:*
```{r}
sd(x)
```

*Range:*
```{r}
#This function returns the highest and lowest value in the vector
range(x)
```

*Summary:*
```{r}
#This funciton outputs a number of useful summary values
summary(x)
```

***

## Plotting Data

The below functions allow you to visualize your data. They generally take one or two vectors as inputs.

*Histogram:*
```{r}
hist(x)
```

*Plot:*
```{r}
#Base plotting function:
plot(y~x,data=data,
     co="color",
     type="p,l")

#col changes the line color
#type changes the type of plot (p = points, l = lines)
```

\hfill\break

*Better plotting:*  

The base plotting function is good for quick-and-dirty plots. But for more complex and publication-quality plots, you’ll want to use the ggplot2 package. This is more complicated, but allows you to create great figures.  

To install a package, use the install.packages() function:
```{r}
install.packages("ggplot2")
```

You only need to do this **once ever**. So don't write this in your script file. Just do it in the console.

Everytime you want to use a package, you need to load it like so:
```{r}
library(ggplot2)
```

\hfill\break

The most basic and most useful function in ggplot2 is the ggplot() function:
```{r}
plot <- ggplot(data = data,
               aes(x = x, y = y)) +
  geom_point()
```

ggplot2 allows you to build a plot by adding options on top of a more basic plot. For example:
```{r}
#geom_smooth() adds a best-fit line to a plot
plot <- ggplot(data = data,
               aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

Other types of plots are possible too:
```{r}
plot <- ggplot(data = data,
               aes(x = x, y = y)) +
  geom_bar(stat = "summary",
           fun = "mean")

```


***

## Data Wrangling

R is really good at analyzing data. It’s also really useful for manipulating data: combining, labeling, rearranging, repeating, etc. Below are a number of functions commonly used to wrangle data.

*Generate a Sequence of Data:*


```{r}

#All values between 1 and 10:

1:10

#Or

seq(1, 10, 1)

#All values between 1 and 10, skipping by 2:
seq(1, 10, 2)
```


*Repeat a Sequence of Data:*

```{r}
#Repeat the number "1" 5 times:
rep(1,times=5)

#Repeat each element of a vector twice:
rep(c(1,2),each=2)

#Repeat a vector twice, while also repeating each element twice:
rep(c(0,1),times=2,each=2)
```


*Put vectors or dataframes together:*

```{r}
#Smoosh two vectors or dataframes together side-by-side (by columns):
cbind(vector1,vector2)

#Smoosh two vectors or dataframes together top-to-bottom (by rows):
rbind(vector1,vector2)

```


*Pick a random subset of data:*
```{r}
#Draw a random sample from data, of size samplesize:
sample(data,samplesize)

#Sample can also be used to generate random data. 
#For example, this will generate a vector of 5 values, randomly equal to 1 or 2:
sample(c(1,2),5,replace=TRUE)
#(replace=TRUE means sample with replacement)

```

***

## Control Flow

R also contains a number of functions that allow you to control how R programs progress with conditional loops and branches.

*If Function:*
```{r}
#An if function executes a section of code only if some statement is true
if(statement) {
  code
}

#For example:

if(4>3) {
  print("Yep!")
}
```

Note that maintaining white space and indentation is very important for keeping if() and similar functions readable. To automatically format an if statement (or any chunk of R code), select the code and hit CTRL + Shift + A on a PC or CMD + Shift + A on a Mac

*Else Function:*
```{r}
#An else function executes a section of code if a statement is false
if(statement) {
  code1
} else {
  code2
}

#For example:
if(3>4) {
  print('Yep! ')
} else{
  print('Nope! ')
}

```

*Ifelse Function:*
```{r}
#An ifelse function crams an if and else function into one line. 
#Very handy! But only if you only need a scalar or vector output.

ifelse(statement, codeIfTrue, codeIfFalse)
ifelse(4 > 3, 'Yep! ', 'Nope! ')
ifelse(3 > 4, 'Yep! ', 'Nope! ')

```

*For Loop:*
```{r}
#A for loop repeats a section of code for a pre-specified number of iterations.

for(i in 1:iterations) {
  code
}

#The “i” acts as a counter that updates itself after each iteration. 
#You can replace “i” with any letter or valid data structure name that you like. 
#Try to make it something that makes sense to you.

#You can use this to, for example, fill a vector step by step:
vector <- rep(0, 100)
for (step in 1:100) {
  vector[step] <- step
}

```

*While Loop:*
```{r}
#A while loop repeats a section of code so long as a statement is true.
while(statement) {
  code
}

#For example:
a <- 0
b <- 1000
while (sum(a) < b) {
  #This will keep appending values to a until the sum is greater than b
  a <- c(a, length(a))
}

```

*Apply Function:*
```{r}

#An apply function repeats a function over each row and column of a dataframe.
apply(data,row/col,function(x) 
  function
  )
#Data refers to the data you want to manipulate, 
#row/col refers to whether you want the function to repeat across rows or cols
#(1 for rows) or columns (2 for columns), 
#function refers to the function you want repeated. 
#You can replace “x” with whatever data structure name you like.

#For exmaple, this takes the sum of every row. Equivalent to rowSums(data):
apply(data,1,function(x) 
  sum(x)
  )

```

*Other Apply Functions:*
```{r}
#sapply applies a function to each element of a vector
sapply(vector,function(x)
  function
  )

#lapply applies a function to vector1 separately for each level of vector 2
lapply(vector1,vector2,function(x)
  function
  )

```

***

## Custom Functions

Finally, if all else fails and R simply doesn’t have a function to do what you want to do, you can write your own custom function! This function can take whatever inputs you desire, process those inputs however you want, and output whatever you like.

```{r}

#To define a custom function, use the function() function:

functionName <- function(inputs) {
  code
  return(output)
}

#To use your function, run the function definition code
#Then call your function like any other:
functionName(inputs)

#For example:
addTwoValues <- function(value1, value2) {
  output <- value1 + value2
  return(output)
}
addTwoValues(1, 2)


```

\newpage
# Cheat Sheet

Action                   | Code 
-------------------------|---------------------------------------------------
Assignment               | <-
Indexing                 | structureName[row,col]; data\$variable
Install packages         | install.packages("packageName")
Load package             | library(package)
Import data              | data<-read.csv(file.choose())
Save data                | write.csv(data,file.choose(),row.names=F)
-------------------------|---------------------------------------------------
Arithmetic operators     | +, -, *, /, ^, sqrt(x), sum(x)
Comparisons and logicals | == ,!=, >, >=, <, <=, %in%, &, \|
Length and sample size   | length(vector), nrow(dataframe)
Mean                     | mean(vector,na.rm=T),colMeans(data,na.rm=T)
Median                   | median(vector)
Standard deviation       | sd(vector)
Range                    | range(vector)
Summary                  | summary(vector)
Single sample t-test     | t1<-t.test(data\$y,mu)
Between-subjects t-test  | t1<-t.test(y~x,data=data)
Correlation              | cor(data\$x,data\$y)
Regression               | r1<-lm(y~x1+x2,data=data)
One-way ANOVA            | a1<-lm(y~x,data=data)
Two-way ANOVA            | a1<-lm(y~x1*x2,data=data)
Repated Measures ANOVA   | a1<-aov(y~x+Error(PIN),data=data)
-------------------------|---------------------------------------------------
Histogram                | hist(vector)
Basic plot               | plot(y~x,data=data)
Ggplot2 plot             | ggplot(data=data,aes(x,y,color=z))+geom_point()
-------------------------|---------------------------------------------------
Generate a sequence      | 1:10, seq(start,finish,by=skip)
Repeat data              | rep(data,times=t,each=e)
Merge data structures    | cbind(vector1,vector2), rbind(vector1,vector2)
Sample from data         | sample(data,size,replace=T/F)
Random binomial data     | rbinom(ncoins,nflips,prob)
Random uniform data      | runif(ndice,min,max)
Random normal data       | rnorm(n,mean,sd)
-------------------------|----------------------------------------------------
If function              | if(statement){code}
Else function            | if(statement){code1} else {code2}
Ifelse function          | ifelse(statement,code1,code2)
For loop                 | for(i in 1:iterations){code}
While loop               | while(statement){code}
Apply function           | apply(data,row/column,function(x) function)
sApply function          | sapply(vector,function(x) function)
lApply function          | lapply(vector1,vector2,function(x) function)
Custom function          | functionName<-function(inputs){code return(output)}

